Continuous risk-based predictive approaches to treatment effect heterogeneity: A simulation study

### 2021-06-28 submission/manuscript.tex
update figures and rebuild
changes in discussion after DvK comments

\begin{abstract}
\textbf{Objective:} Compare risk-based methods for individualizing
treatment effects with simulations in the RCT setting. \textbf{Study
Design and Setting:} We predicted absolute benefit based on an available
prognostic index (PI) using: a model with the PI and a constant relative
treatment effect; a model including an interaction of treatment with the
PI; 4 quarters of the PI; nonlinear transformations of the PI
(restricted cubic splines with 3, 4 and 5 knots); an adaptive model
selection method using Akaike's Information Criterion. Starting from a
base case scenario (sample size 4,250, constant odds ratio 0.8, AUC of
the PI 0.75), we considered diverse assumptions by introducing linear
and quadratic interactions of the PI with treatment and varying sample
size and discriminative ability of the PI. We evaluated performance
using root mean squared error, discrimination and calibration for
benefit. \textbf{Results:} Models including a linear interaction of the
PI with treatment had adequate performance that was robust under most
simulation scenarios. Restricted cubic splines required larger sample
sizes and higher AUC of the PI to achieve adequate performance. The
adaptive approach performed equivalently to the best-performing method
in each scenario. \textbf{Conclusion:} Usually, models with just a
linear interaction of the PI with treatment adequately predict absolute
benefit.
\end{abstract}

@@

Within the setting of patient-centered outcomes research, predictive
approaches for assessing heterogeneity of treatment effects (HTE) aim at
the development of models predicting either individualized effects or
which of two (or more) treatments is better for an individual {[}1{]}.
In prior work, we divided such methods in three broader categories based
on the reference class used for defining patient similarity when making
individualized predictions or recommendations {[}2{]}. Risk-modeling
approaches use prediction of baseline risk as the reference; treatment
effect modeling approaches also model treatment-covariate interactions,
in addition to risk factors; optimal treatment regime approaches focus
on developing treatment assignment rules and therefore rely heavily on
modeling treatment effect modifiers.

Risk-modeling approaches to predictive HTE analyses provide a viable
option in the absence of well-established treatment effect modifiers
{[}3,4{]}. In simulations, modeling of effect modifiers in the form of
treatment-covariate interactions often led to miscalibrated predictions
of benefit, while risk-based methods proved quite robust {[}5{]}. Most
often, risk-modeling approaches are carried out in two steps: first a
risk prediction model is developed externally or internally on the
entire RCT population, ``blinded'' to treatment; then the RCT population
is stratified using this prediction model to evaluate risk-based
treatment effect variation {[}6{]}. However, even though estimates at
the risk subgroup level are accurate, this does not apply on the
individual level, especially for patients with predicted risk at the
boundaries of the risk intervals. Therefore, the risk-stratified
approach should be used for exploring and presenting an overview of HTE,
while inferences on the individual level should be made with caution.

We aimed to provide an overview of methods that can be used to move from
a risk-stratified approach to a continuous one using common smoothing
techniques. These methods extend the risk-based framework of predictive
HTE analyses to allow predictions on the individual level, within the
RCT setting. We carried out a simulation study to compare the
performance of these methods under different settings of increasing
non-linearity of treatment effects. Finally, we carried out an
application on real data as a demonstration of the considered
techniques.

@@

The linear interaction model proved to be flexible enough to adequately
predict benefit under a wide range of scenarios. It had lower error
rates and usually the highest c-for-benefit with the lower variability
across simulation replications. Its benefit predictions were also
well-calibrated across the majority of the simulation settings. Models
with RCS smoothing with 3 knots performed very similarly, only
outperforming the linear interaction model in the presence of strong
quadratic deviations from the base case scenario of constant relative
treatment effects. These simpler methods proved to have quite robust
performance across all scenarios, which was not the case for more
flexible approaches.

Methods with RCS smoothing with 4 or 5 knots proved to be data-hungry
and this often led to overfitting with smaller sample sizes. This
resulted in higher RMSE, increased variability of discrimination for
benefit and worse-calibrated benefit predictions across simulation
replications. Even with larger sample sizes and strong quadratic
deviations from the base case scenario of constant relative treatment
effects, these more flexible RCS smoothing methods did not outperform
simpler RCS smoothing with 3 knots. More flexible approaches usually do
not come with any added value, unless we expect extreme heterogeneity in
treatment effects.

The constant treatment effect model, despite having adequate performance
even in the presence of weak treatment effect heterogeneity on the
relative scale, it quickly broke down with stronger deviations from the
base case scenario of constant effects. In these cases, the stratified
approach generally had lower error rates compared to the constant
treatment effect model. In concept, the stratified approach lies between
the constant effects model and smoother approaches, only assuming
constant treatment effects within strata of predicted risk and,
therefore, is more sensitive to treatment effect heterogeneity. However,
these stepwise benefit estimates can only be used as a demonstration of
treatment effect heterogeneity, identifying patient subgroups where
treatment benefits are expected to be higher or lower, and should not be
used to make individualized absolute benefit predictions.

Increasing the true prediction AUC by increasing the predictor
coefficients of the true risk model reduced RMSE for all methods,
similar to increasing the sample size. Higher AUC translates in higher
variability of predicted risks, which, in turn, allows the considered
methods to better follow the evolution of treatment benefit. Higher
prediction AUC also led to higher absolute values of c-for-benefit. Even
though, model performance appears to be very important for the
performance of these methods, prediction model development was outside
the scope of this work and has already been studied quite extensively
{[}5,12,13{]}.

Risk-based approaches to predictive HTE assume that benefit is a
function of baseline risk. A limitation of our study is that this was
assumed to be the case in the majority of the simulation scenarios. Even
though our main conclusions did not change when simulating true
treatment-covariate interactions (results in supplement), RMSE had
higher values. The linear interaction model under these scenarios again
had the best overall performance under different sizes of the
interaction effects. This can be explained by the fact that linear
interactions with treatment were simulated, thus favoring the linear
interaction model. In studies evaluating prediction model development,
the underlying outcome generation process sometimes proved to affect
performance of certain methods {[}14{]}. This might be the case here as
well and should be explored in more extensive future simulation studies.

Recent years have seen an increased interest in predictive HTE
approaches focusing on individualized benefit predictions. In our
simulations we only focused on risk-based methods, using baseline risk
as a reference in a two-stage approach to individualizing benefit
predictions. However, there is a plethora of different methods, ranging
from treatment effect modeling to tree-based approaches available in
more recent literature {[}15--17{]}. Simulations are also needed to
assess relative performance and define the settings where these break
down or outperform each other.

In conclusion, when comparing different risk-based approaches to
individualizing benefit predictions simpler models including a linear
interaction with the prognostic index or RCS smoothing with 3 knots had
adequate performance under a wide range scenarios. An adaptive approach
selecting among the considered methods with optimal AIC had satisfactory
performance.

@@

\leavevmode\hypertarget{ref-Austin2021}{}%
{[}14{]} Austin PC, Harrell FE, Steyerberg EW. Predictive performance of
machine and statistical learning methods: Impact of data-generating
processes on external validity in the ``large n, small p'' setting.
Statistical Methods in Medical Research 2021;30:1465--83.
\url{https://doi.org/10.1177/09622802211002867}.

\leavevmode\hypertarget{ref-Athey2019}{}%
{[}15{]} Athey S, Tibshirani J, Wager S. Generalized random forests. The
Annals of Statistics 2019;47. \url{https://doi.org/10.1214/18-aos1709}.

\leavevmode\hypertarget{ref-Lu2018}{}%
{[}16{]} Lu M, Sadiq S, Feaster DJ, Ishwaran H. Estimating individual
treatment effect in observational data using random forest methods.
Journal of Computational and Graphical Statistics 2018;27:209--19.
\url{https://doi.org/10.1080/10618600.2017.1356325}.

\leavevmode\hypertarget{ref-Wager2018}{}%
{[}17{]} Wager S, Athey S. Estimation and inference of heterogeneous
treatment effects using random forests. Journal of the American
Statistical Association 2018;113:1228--42.
\url{https://doi.org/10.1080/01621459.2017.1319839}.
\end{cslreferences}

>>>>

\begin{abstract}
\textbf{Objective:} Baseline risk is an important determinant of the
absolute effect of a treatment for an individual patient. We aimed to
compare different risk-based methods for predicting individualized
treatment effects with simulations of the RCT setting. \textbf{Study
Design} and Setting: We sampled patients from a superpopulation with
diverse assumptions for a baseline prognostic index of risk (PI), and
for the shape of the interaction between the PI and treatment (no,
linear or quadratic interaction). In each sample, we fitted different
models for predicting absolute benefit: a model with the PI and a
constant relative treatment effect and models including an interaction
of treatment assignment with the PI, with 4 quarters of the PI, and with
nonlinear transformations of the PI (restricted cubic splines with 3, 4
and 5 knots). We also used an adaptive model selection approach based on
Akaike's Information Criterion. We evaluated predictive performance in
the superpopulation using root mean squared error and measures of
discrimination and calibration for benefit. Starting from a base case
scenario (sample size 4,250, treatment odds ratio 0.8, AUC of the PI
0.75), we varied the sample size, the strength of the treatment effect,
and the discriminative ability of the PI. \textbf{Results:} Models
including a linear interaction of the PI with treatment had (close to?)
optimal performance under most simulation scenarios. More flexible
models with restricted cubic splines required larger sample sizes and
higher AUC of the PI to outperform the linear model. The adaptive
approach performed similarly to the best-performing method in each
scenario. \textbf{Conclusion:} Under most circumstances, a model with a
linear interaction of the PI with treatment is better able to predict
absolute treatment benefit.
\end{abstract}

@@

Predictive approaches for assessing heterogeneity of treatment effects
(HTE) aim at the development of models predicting either individualized
effects or which of two (or more) treatments is better for an individual
{[}1{]}. In prior work, we divided such methods in three broader
categories based on the reference class used for defining patient
similarity when making individualized predictions or recommendations
{[}2{]}. Risk-modeling approaches use prediction of baseline risk as the
reference; treatment effect modeling approaches also model
treatment-covariate interactions, in addition to risk factors; optimal
treatment regime approaches focus on developing treatment assignment
rules and therefore rely heavily on modeling treatment effect modifiers.

Risk-modeling approaches to predictive HTE analyses provide a viable
option in the absence of well-established treatment effect modifiers
{[}3,4{]}. In simulations, modeling of effect modifiers, i.e.
treatment-covariate interactions, often led to miscalibrated predictions
of benefit, while risk-based methods proved quite robust {[}5{]}. Most
often, risk-modeling approaches are carried out in two steps: first a
risk prediction model is developed externally or internally on the
entire RCT population, ``blinded'' to treatment; then the RCT population
is stratified using this prediction model to evaluate risk-based
treatment effect variation {[}6{]}. However, even though estimates at
the risk subgroup level may be accurate, these estimated do not apply to
individual patients, especially for patients with predicted risk at the
boundaries of the risk intervals. Hence, the risk-stratified approach is
useful for exploring and presenting HTE, but is not useful for
supporting treatment decisions for individual patients.

To individualize treatment effects, the recent PATH statement suggested
various risk-based models including a prognostic index of baseline risk
(PI) and treatment assignment {[}cite PATH !!{]}. We aimed to summarize
and compare different risk-based models for predicting individualized
treatment effects. We simulated RCT settings to compare the performance
of these models under different assumptions of the relationship between
baseline risk and treatment. We illustrated the different models by a
case study of predicting individualized effects of tissue plasminogen
activator (tPA) versus streptokinase treatment in patients with an acute
myocardial infarction (MI).

@@

The linear interaction model proved to be superior to alternative
approaches when predicting risk-based treatment benefit under a wide
range of scenarios. It generally had lower mean squared error, lower
e-for-benefit and higher c-for-benefit with lower variability across
simulation replications. Models including restricted cubic splines with
3 knots only outperformed the linear interaction model in the presence
of strong quadratic deviations from a constant relative treatment
effect.

Including restricted cubic splines with 4 or 5 knots proved to be too
flexible, as indicated by higher RMSE, increased variability of
discrimination for benefit and worse calibration of benefit predictions.
Even with larger sample sizes and strong quadratic deviations from the
base case scenario of constant relative treatment effects, these more
flexible restricted cubic splines did not outperform the simpler RCS
with 3 knots. These more flexible approaches may only be helpful if we
expect more extreme patterns of heterogeneous treatment effects.

The constant treatment effect model, despite having adequate performance
in the presence of weak treatment effect heterogeneity on the relative
scale, quickly broke down with stronger deviations from constant
relative treatment effects. In these cases, the stratified approach
generally had lower error rates compared to the constant treatment
effect model. In concept, the stratified approach lies between the
constant effects model and smoother approaches, only assuming constant
treatment effects within strata of predicted risk and, therefore, is
more sensitive to treatment effect heterogeneity. Stepwise treatment
benefit estimates are very useful for demonstrating treatment effect
heterogeneity -- because estimating treatment effect requires groups of
patients rather than individual patients --but are not helpful for
making individualized absolute benefit predictions.

Increasing the discriminative ability of the risk model -- by increasing
the predictor coefficients of the true risk model -- reduced RMSE for
all methods, similar to increasing the sample size. This increase in
discriminative ability translates in higher variability of predicted
risks, which, in turn, allows the considered methods to better follow
the evolution of treatment benefit. As expected, the increase in
discriminative ability of the risk model also led to higher values of
c-for-benefit. Even though risk model performance is very important for
the ability of risk-based methods to predict treatment benefit,
prediction model development was outside the scope of this work and has
already been studied extensively {[}5,12,13{]}.

Risk-based approaches to predictive HTE estimate treatment benefit as a
function of baseline risk. A limitation of our study is that we assumed
treatment benefit to be a function of baseline risk in the majority of
the simulation scenarios. Nevertheless, our main conclusions did not
change when we generated individual treatment-covariate interactions
(supplementary table/figure x). Future simulation studies could explore
the effect of more extensive deviations from risk-based treatment
effects.

Recent years have seen an increased interest in predictive HTE
approaches focusing on individualized benefit predictions. In our
simulations we only focused on risk-based methods, using baseline risk
as a reference in a two-stage approach to individualizing benefit
predictions. However, there is a plethora of different methods, ranging
from treatment effect modeling to tree-based approaches available in
more recent literature {[}14--16{]}. Simulations are also needed to
assess relative performance and define the settings where these break
down or outperform each other.

In conclusion, when comparing different risk-based approaches to predict
individualizing benefit a models including a linear treatment
interaction with the prognostic performed best in a wide range of
scenarios. More flexible models with restricted cubic splines required
larger sample sizes and higher AUC of the PI to outperform the linear
model. An adaptive approach, selecting the model with optimal AIC, had
satisfactory performance.

@@

\leavevmode\hypertarget{ref-Athey2019}{}%
{[}14{]} Athey S, Tibshirani J, Wager S. Generalized random forests. The
Annals of Statistics 2019;47. \url{https://doi.org/10.1214/18-aos1709}.

\leavevmode\hypertarget{ref-Lu2018}{}%
{[}15{]} Lu M, Sadiq S, Feaster DJ, Ishwaran H. Estimating individual
treatment effect in observational data using random forest methods.
Journal of Computational and Graphical Statistics 2018;27:209--19.
\url{https://doi.org/10.1080/10618600.2017.1356325}.

\leavevmode\hypertarget{ref-Wager2018}{}%
{[}16{]} Wager S, Athey S. Estimation and inference of heterogeneous
treatment effects using random forests. Journal of the American
Statistical Association 2018;113:1228--42.
\url{https://doi.org/10.1080/01621459.2017.1319839}.
\end{cslreferences}
